Think deep for reef resilience

### 2025-08-22 paper/paper_draft.qmd
Add two-way interaction PDP plot to paper and move CID table to sup

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Our findings suggest that depth provides critical
    protection against thermal stress, contrasting current restoration practices that primarily
    target easily accessible shallow sites. These results have important implications for
    reef management, indicating that depth refugia should be prioritized for further
    consideration in conservation planning alongside connectivity considerations to maximize
    coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
editor:
  markdown:
    wrap: 92
---

@@

Of the GCMs explored, NorESM2-MM had the lowest Equilibrium Climate Sensitivity (2.5°C) and
reefs exhibit continuing recovery and decline dynamics up to 2075 under this hypothetical
climate future. Unsurprisingly, recovery phases were predominantly seen in the reefs in the
High cluster, with Medium and Low exhibiting lesser probability of such periods of
recovery. Generally, recovery and increases in coral cover were visible in simulated results
up to 2045. Under NorESM2-MM, the Far Northern Management area were found to have the most
similar temporal behaviour between Low and High clusters compared to other Management
areas (@tbl-cluster-cid) due to the low degree of coral cover overall
(@fig-gcm-cover-trajectories).

Intermediate levels of coral cover maintenance can be seen in ACCESS-ESM1-5 (ECS value of
3.87°C) when compared to the other GCMs (of higher and lower ECS values). Under these
intermediate conditions (in relative terms), there is potential for reefs in the High
cluster to maintain coral cover levels around 20% to 2050. The highest peak DHW levels
occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHWs time
series occur under ACCESS-CM2 (@fig-methods-dhws). Under conditions represented by high ECS
value GCMs - specifically, ACCESS-CM2 (4.66°C), EC-Earth3-Veg (4.30°C) and GFDL-CM4
(4.1°C) - all reefs are largely in decline across all management areas from 2055
(@tbl-cluster-cid).

@@

```{julia}
#| echo: false
#| warning: false
#| message: false
#| results: asis
#| label: tbl-cluster-cid
#| tbl-cap: Complexity invariant distance (CID) between low and high cluster median time series in each management area under each GCM. CID values are calculated for percentage cover time series (@fig-gcm-cover-trajectories) and Degree Heating Week (DHW) time series (@sfig-gcm-dhw-trajectories). Complexity invariant distance represents the Euclidean distance between time series, corrected by the complexity of the two time series. CID values are used to quantify overall differences between these two cluster levels at management area and GCM levels. CID values for percentage cover and DHW time series are not directly comparable as the time series use different units.

@@

@@

Under all GCMs many reefs are projected to suffer substantial declines in coral cover
relative to 2025 levels, with an average of 38% of reefs (924 reefs) clustered into the
lowest cover level cluster at the management area scale across GCMs. NorESM2-MM and
ACCESS-ESM1-5 GCMs convey the most coral cover maintenance into the future. Differences
between Low, Medium and High reef clusters in cover time series are most notable under these
two GCMs (@tbl-cluster-cid; @fig-gcm-cover-trajectories). Maintenance of coral cover into
the future is linked to the Equilibrium Climate Sensitivity of the GCM used as levels of
future recovery and timespan of cover maintenance increase as GCM ECS increases
(@fig-gcm-cover-trajectories). Under high-ECS GCMs (GFDL-CM4, EC-Earth3-Veg and ACCESS-CM2)
large declines and lower rates of recovery in high cover clusters are common compared to
lower-ECS GCMs.

@@

>>>>

---
title: Think deep for reef resilience
author:
  - name: Benjamin Grier
    email: grierben777@gmail.com
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          state: QLD
    attributes:
        corresponding: true
  - name: Takuya Iwanaga
    email: t.iwanaga@aims.gov.au
    affiliations:
        - id: AIMS
          name: Australian Institute of Marine Science
          address: Cape Cleveland Rd
          city: Cape Cleveland
          state: QLD
    attributes:
        corresponding: true
abstract: |
    Coral reefs globally face unprecedented threats from climate change, with the Great Barrier
    Reef (GBR) experiencing severe declines in coral cover from thermal stress events.
    Understanding drivers behind reef resilience to climate impacts is critical for conservation
    planning and intervention strategies. A scenario-based population modelling approach was
    adopted with larval connectivity dynamics, thermal adaptation, and environmental factors
    to assess coral reef resilience across GBR locations under future climate projections.
    Using five Global Climate Models (GCMs) spanning five Equilibrium Climate Sensitivities,
    we analysed reef cover time series and the ability of reefs to maintain positive carbonate
    production budgets under future conditions. Contrary to expectations, larval connectivity
    patterns did not correlate with increases in maintenance of positive carbonate budgets.
    Instead, reef depth emerged as the primary predictor of cover-time series resilience and
    maintenance of positive carbonate budgets, with deeper reefs benefiting from thermal
    refugia that reduce bleaching impacts. Our findings suggest that depth provides critical
    protection against thermal stress, contrasting current restoration practices that primarily
    target easily accessible shallow sites. These results have important implications for
    reef management, indicating that depth refugia should be prioritized for further
    consideration in conservation planning alongside connectivity considerations to maximize
    coral survival under continuing climate change.
keywords:
  - Coral reef
  - Climate change
  - Ecosystem resilience
  - Reef depth
  - Reef connectivity
date: last-modified
format:
  elsevier-docx:
    journal:
      name: Journal Name
      formatting: preprint
      cite-style: authoryear
bibliography: references.bib
engine: julia
crossref:
  custom:
  - kind: float
    reference-prefix: SFigure
    key: sfig
  custom:
  - kind: float
    reference-prefix: STable
    key: stbl
editor:
  markdown:
    wrap: 92
---

@@

Of the GCMs explored, NorESM2-MM had the lowest Equilibrium Climate Sensitivity (2.5°C) and
reefs exhibit continuing recovery and decline dynamics up to 2075 under this hypothetical
climate future. Unsurprisingly, recovery phases were predominantly seen in the reefs in the
High cluster, with Medium and Low exhibiting lesser probability of such periods of
recovery. Generally, recovery and increases in coral cover were visible in simulated results
up to 2045. Under NorESM2-MM, the Far Northern Management area were found to have the most
similar temporal behaviour between Low and High clusters compared to other Management
areas (@stbl-cluster-cid) due to the low degree of coral cover overall
(@fig-gcm-cover-trajectories).

Intermediate levels of coral cover maintenance can be seen in ACCESS-ESM1-5 (ECS value of
3.87°C) when compared to the other GCMs (of higher and lower ECS values). Under these
intermediate conditions (in relative terms), there is potential for reefs in the High
cluster to maintain coral cover levels around 20% to 2050. The highest peak DHW levels
occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHWs time
series occur under ACCESS-CM2 (@fig-methods-dhws). Under conditions represented by high ECS
value GCMs - specifically, ACCESS-CM2 (4.66°C), EC-Earth3-Veg (4.30°C) and GFDL-CM4
(4.1°C) - all reefs are largely in decline across all management areas from 2055
(@stbl-cluster-cid).

@@

@@

![Partial dependence plot indicating the probability of cluster assignment along interacting values of median reef depth (metres) and Log10 weighted incoming connectivity. Colour indicates the probability of assignment to each cluster, with lighter colours indicating higher probability of assignment. Panel A shows the probability of assignment for the Low cover cluster, Panel B shows Medium cover cluster and Panel C shows assignment probability for the High cluster category. Probability values for each cluster are generated by the random forest classifier using each combination of depth and incoming connectivity values along the axes at 50 grid points. The actual distributions of depth and incoming connectivity values are displayed in panels D and E respectively.](../figs/ts_cluster_rf_twoway_depth_conn_pdp.png){#fig-ts-cluster-pdp-twoway fig-align="center"
width="14.82cm"}

@@

Under all GCMs many reefs are projected to suffer substantial declines in coral cover
relative to 2025 levels, with an average of 38% of reefs (924 reefs) clustered into the
lowest cover level cluster at the management area scale across GCMs. NorESM2-MM and
ACCESS-ESM1-5 GCMs convey the most coral cover maintenance into the future. Differences
between Low, Medium and High reef clusters in cover time series are most notable under these
two GCMs (@stbl-cluster-cid; @fig-gcm-cover-trajectories). Maintenance of coral cover into
the future is linked to the Equilibrium Climate Sensitivity of the GCM used as levels of
future recovery and timespan of cover maintenance increase as GCM ECS increases
(@fig-gcm-cover-trajectories). Under high-ECS GCMs (GFDL-CM4, EC-Earth3-Veg and ACCESS-CM2)
large declines and lower rates of recovery in high cover clusters are common compared to
lower-ECS GCMs.

@@

```{julia}
#| echo: false
#| warning: false
#| message: false
#| results: asis
#| label: stbl-cluster-cid
#| tbl-cap: Complexity invariant distance (CID) between low and high cluster median time series in each management area under each GCM. CID values are calculated for percentage cover time series (@fig-gcm-cover-trajectories) and Degree Heating Week (DHW) time series (@sfig-gcm-dhw-trajectories). Complexity invariant distance represents the Euclidean distance between time series, corrected by the complexity of the two time series. CID values are used to quantify overall differences between these two cluster levels at management area and GCM levels. CID values for percentage cover and DHW time series are not directly comparable as the time series use different units.
